---
title: "exploratory data analysis"
author: "Peter"
date: '2019-01-02'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Dependencies

```{r load dependencies}

library(tidyverse)
library(DataExplorer)
```


## Read the data

I will analyse the training data set to learn more about the data itself.

```{r read data}

train <- read.csv("../../data/adult.data", 
                  header = FALSE,
                  stringsAsFactors = TRUE)

test <- read.csv("../../data/adult.test", 
                  header = FALSE,
                  stringsAsFactors = TRUE)
```

## Data cleaning

```{r data cleaning}

# add column names into the data frame

column_name_string <- c("age", "workclass", "fnlwgt", "education", "education_num", "marital_stat", "occupation", "relationship", "race", "sex", "capital_gain", "capital_loss", "hr_per_wk", "native_country", "label")

# ref: https://stackoverflow.com/questions/6081439/changing-column-names-of-a-data-frame
colnames(train) <- column_name_string
colnames(test) <- column_name_string
```

## Exploratory Data Analysis

*Examine the variables*

I try not to look at any pre-exisitng tutorials so I am not biased towards any particular analysis approach. However, the variable `fnlwgt` was not explained by the original data source so I did a quick google search. I found a short answer on Quora (link: https://www.quora.com/What-is-the-meaning-of-capital-gain-capital-loss-and-fnlwgt-in-adult-dataset-from-UCI). `fnlwgt` means sampling weight. 

A tutorial by Statistical Consulting Group of (http://scg.sdsu.edu/dataset-adult_r/) on this data set suggested ignoring the `fnlwgt` variable because it clutters the analysis. `fnlwgt` represents the number of people the census takers believe that observation represents. This sampling weight also seems unnecessary for the purpose of my analysis because my analysis is a proof of concept. This data set may also be too small to make a proper inference of the whole US population. 

Secondly, some variables seem to be highly correlated. level of education `education` and years of eduation `education_num` pretty much represent the same idea. `relationship` and `marital_status` are also highly correlated. The statistical consulting group also points out that `relationship` contains family relationship values such husband, father, and so on, but it only contains one per observation. Some information is missing from this `relationship` variable because a person can have multiple relationships if it is a big family in the household. Also, having too many features may result in a overfitted classifier. I am attempted to remove `education_num` and `relationship`. (ref: https://datascience.stackexchange.com/questions/24452/in-supervised-learning-why-is-it-bad-to-have-correlated-features)

However, it may be too early to remove highly correlated features at this stage because I do not have a lot of domain knowledge on census data and donot predictors. I will keep these variables in mind when I perform feature selections at a later stage. (ref: https://datascience.stackexchange.com/questions/36404/when-to-remove-correlated-variables) 

Note to self: These three variables should be re-examined for its relevance as a feature 
- `fnlwgt` 
- `education_num`
- `relationship` 

*Data Set Summary*

```{r summary1}

introduce(train)


```


```{r summary2}

introduce(test)

```


There are 14 features (+ 1 label = 15 columns) and 32561 rows of observations. Professor Mark Schmidt (UBC CS) suggested that the data should be at least ten times the number of features so we have enough observations to build a model. 

Initial findings suggest no missing columns and missing values are present in the training and test sets. 

*Examine categorical variables*

```{r workclass}

plot_bar(train$workclass, title = "work class")

```

```{r education}

plot_bar(train$education, title = "education")

```


```{r marital status}

plot_bar(train$marital_stat, title = "marital status")

```


```{r Occupation}

plot_bar(train$occupation, title = "occupation")

```

```{r relationship}

plot_bar(train$relationship, title = "relationship")

```


```{r race}

plot_bar(train$race, title = "race")

```

```{r sex}

plot_bar(train$sex, title = "gender")

```


```{r native country}

plot_bar(train$native_country, title = "native countries")

```


```{r label}

plot_bar(train$label, title = "label")

```


My previous finding did not detect any missing values. It turns out that some `?` values are present in the `workclass`, `occupation`, and `native_country` feature. I will treat these values as missing data and remove them from training and test set.

*Remove missing data from categorical variables*

(code ref: https://suzan.rbind.io/2018/02/dplyr-tutorial-3/#filtering-based-on-multiple-conditions)

```{r count missing data in training set}

# count missing data in training set
train %>% 
  filter(str_detect(workclass, "\\?") | str_detect(occupation, "\\?") | str_detect(native_country, "\\?")) %>% 
  count()


```

```{r count missing data in test set}

# count missing data in test set
test %>% 
  filter(str_detect(workclass, "\\?") | str_detect(occupation, "\\?") | str_detect(native_country, "\\?")) %>% 
  count()
```

I need to remove 2399 rows from training set and 1221 rows from test set. I am removing less than 10% of the observations so I am not concerned.

```{r remove missing data}

# remove missing values from training set
train_completed <- train %>% 
  filter(!str_detect(workclass, "\\?") & !str_detect(occupation, "\\?") & !str_detect(native_country, "\\?"))

# remove missing values from test set
test_completed <- test %>% 
  filter(!str_detect(workclass, "\\?") & !str_detect(occupation, "\\?") & !str_detect(native_country, "\\?"))
  
```

```{r verify data set integrity}

# check and see if I have removed all rows with missing values
# the sum of rows should equal 32561 for training and 16282 for test

# check training set
nrow(train) == (nrow(train_completed) + nrow(train %>% 
  filter(str_detect(workclass, "\\?") | str_detect(occupation, "\\?") | str_detect(native_country, "\\?"))))

```

```{r verifty data set integrity2}

# check test set
nrow(test) == (nrow(test_completed) + nrow(test %>% 
  filter(str_detect(workclass, "\\?") | str_detect(occupation, "\\?") | str_detect(native_country, "\\?"))))

```

*Examine continuous variables*

```{r age}

plot_histogram(train$age, title = "age distribution") 

```

```{r fnlwgt}

plot_histogram(train$fnlwgt)

```

